Very simplified likelihood

### 2024-09-17 inst/hankin-AustrianJournalofStatistics.tex
corrections

(up to an arbitrary additive constant).  Using
$\mathcal{S}^*=\log(B-1)+B\log(1-r/n)$ as an approximate support thus
incurs an error of only $\mathcal{O}(n^{-1})$.  Observing that
$\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have $\hat{B}=1-\log(1-r/n)^{-1}$, or
alternatively $\hat{a}=(1-\log(1-r/n))^{-1}$.  Compare this with the
result of asking ``what is the probability that the focal competitor
beats a randomly chosen opponent?'' to which the binomial maximum
likelihood estimate is $\frac{n-r}{n}$.  For repeated independent
observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have

\begin{equation}
  \hat{a} =   \frac{N}{N+\sum\log(1-r_i/n_i)};
\end{equation}

compare the binomial result of $1-\left.\sum r_i\right/\sum n_i$.
Below, I present three examples that showcase likelihood
function~\ref{likeforrn3}, drawn from the fields of athletics,
education, and Formula~1 motor racing.

@@

\subsection{Olympic track and field athletics}

One long-standing contention in the athletic world is that of lane
bias~\citep{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Men's 100m Olympic running from
2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
competitors' ranks are used.  Figure~\ref{plotolymp} shows a
log-likelihood curve for the Bradley--Terry strength of the focal
competitor, here the runner in the innermost lane.  We see that the
evaluate of $0.474$ is very close to the neutral value of $0.5$, for
which the support is only $0.16$, far short of the standard two units
of support criterion for rejection~\citep{edwards1972}.  This dataset
contains no evidence to support lane bias.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
\caption{Log-likelihood curve for the Bradley--Terry strength of the
  lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
\end{centering}
\end{figure}

@@

``Parkrun'' is a distributed community initiative that organises
weekly timed 5 km runs/walks in parks worldwide~\citep{hindley2020}.
A typical event will have 200 participants.  Table~\ref{parkruntable}
shows the author has completed a total of 21 parkruns to date,
comprising 9 runs in Auckland and 12 at Stirling.  From the first pair
of numbers we see that 259 runners attended that particular parkrun at
Auckland, of whom the author placed 177.

\begin{table}[t]
  \centering
  \caption{Parkrun results}
\label{parkruntable}
\begin{tabular}{cccccccccc}\\
 & \multicolumn{9}{c}{Auckland}\\
rank   & 177& 222& 206& 142& 118& 224& 128& 115& 183\\
runners& 259& 305& 297& 241& 179& 338& 203& 245& 254\\ \\
\end{tabular}
\begin{tabular}{ccccccccccccc}
  & \multicolumn{12}{c}{Stirling}\\
rank   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
runners&  145& 157& 156& 191& 172& 174& 173& 165& 172& 199& 181& 229\\
\end{tabular}
\end{table}

Is there evidence that the author's performance differs between
Auckland and Stirling?  Figure~\ref{parkruncontour} shows contours of
support [again normalised so the support of the evaluate is zero] for
the parkrun dataset, as a function of Auckland strength on the
horizontal axis and Stirling strength on the vertical axis.  A null of
equal strengths appears as the diagonal line.  We see that the maximum
likelihood estimate is $\hat{p}_\mathrm{Auck}=0.427$,
$\hat{p}_\mathrm{Stir}=0.408$.  However, a constrained optimization
shows that the maximum support along the null diagonal is about
$-2.43$, at ${p}_\mathrm{Auck}={p}_\mathrm{Stir}\simeq 0.409$.  The
two units of support criterion~\citep{edwards1972} is met, and we are
justified in concluding that the student is indeed more competent at
applied than pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{parkruncontour} % produced by very_simplified_likelihood.Rmd
\caption{Support contours for Bradley-Terry strength
  at \label{parkruncontour} Auckland and Stirling.  Null of equal
  strengths shown as a diagonal line}
\end{centering}
\end{figure}

@@

\begin{table*}[t]
  \caption{Educational dataset}
\label{educationtable}
\begin{tabular}{llll}
\hline
                   course&category   &rank&class size\\ \hline
      rings and modules  &  pure     & 9  &      12\\
           group theory  &  pure     & 7  &      17\\
               calculus  & applied   & 2  &      23\\
         linear algebra  & applied   & 3  &       9\\
 differential equations  & applied   & 2  &      13\\
               topology  &   pure    & 8  &      14\\
     special relativity  & applied   & 5  &      13\\
        fluid mechanics  & applied   & 4  &      12\\
            Lie algebra  &   pure    & 9  &      15\\
     \hline
\end{tabular}
\end{table*}

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$, at
${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.713$.  The two units of
support criterion~\citep{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

>>>>

(up to an arbitrary additive constant).  Using
$\mathcal{S}^*=\log(B-1)+B\log(1-r/n)$ as an approximate support thus
incurs an error of only $\mathcal{O}(n^{-1})$.  Observing that
$\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have $\hat{B}=1-\log(1-r/n)^{-1}$, or
alternatively $\hat{a}=(1-\log(1-r/n))^{-1}$.  Compare this with the
result of asking ``what is the probability that the focal competitor
beats a randomly chosen opponent?'' to which the binomial maximum
likelihood estimate is $\frac{r-1}{n-r}$.  For repeated independent
observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have

\begin{equation}
  \hat{a} =   \frac{1}{1-\sum\log(1-r_i/n_i)/N};
\end{equation}

compare the binomial result of $\sum(r_i-1)/\sum n_i$.  Below, I
present three examples that showcase likelihood
function~\ref{likeforrn3}, drawn from the fields of athletics,
education, and Formula~1 motor racing.

@@

% \subsection{Olympic track and field athletics}
% 
% One long-standing contention in the athletic world is that of lane
% bias~\citep{munro2022}.  Running in the inner lane is frequently held
% to be a disadvantage; mechanisms include biomechanical issues arising
% from centrifugal acceleration, to competitors' inability to see their
% rivals.  Here attention is restricted to heats, for which lane placing
% is random.  I consider every heat in Men's 100m Olympic running from
% 2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
% competitors' ranks are used.  Figure~\ref{plotolymp} shows a
% log-likelihood curve for the Bradley--Terry strength of the focal
% competitor, here the runner in the innermost lane.  We see that the
% evaluate of $0.511$ is very close to the neutral value of $0.5$, for
% which the support is only $0.029$, far short of the standard two units
% of support criterion for rejection~\citep{edwards1972}.  This dataset
% contains no evidence to support lane bias.
% 
% \begin{figure}[t]
%   \begin{centering}
% \includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
% \caption{Log-likelihood curve for the Bradley--Terry strength of the
%   lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
% \end{centering}
% \end{figure}

@@

``Parkrun'' is a distributed community initiative that organises
weekly timed 5 km runs/walks in parks worldwide~\citep{hindley2020}.
A typical event will have 200 participants.  Table~\ref{parkruntable}
shows the author has completed a total of 21 parkruns to date,
comprising 9 runs in 2023 and 12 in 2024.  From the first pair of
numbers we see that 259 runners attended that particular parkrun in
2023, of whom the author placed 177.

\begin{table}[t]
  \centering
  \caption{Parkrun results}
\label{parkruntable}
\begin{tabular}{cccccccccc}\\
 & \multicolumn{9}{c}{2023}\\
rank   & 177& 222& 206& 142& 118& 224& 128& 115& 183\\
runners& 259& 305& 297& 241& 179& 338& 203& 245& 254\\ \\
\end{tabular}
\begin{tabular}{ccccccccccccc}
  & \multicolumn{12}{c}{2024}\\
rank   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
runners&  145& 157& 156& 191& 172& 174& 173& 165& 172& 199& 181& 229\\
\end{tabular}
\end{table}

Is there evidence that the author's performance differs between
years?  Figure~\ref{parkruncontour} shows contours of
support [again normalised so the support of the evaluate is zero] for
the parkrun dataset, as a function of 2023 strength on the
horizontal axis and 2024 strength on the vertical axis.  A null of
equal strengths appears as the diagonal line.  We see that the maximum
likelihood estimate is $\hat{p}_\mathrm{2023}=0.427$,
$\hat{p}_\mathrm{2024}=0.408$.  However, a constrained optimization
shows that the maximum support along the null diagonal is about
$-2.43$, at ${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.409$.  The
two units of support criterion~\citep{edwards1972} is met, and we are
justified in concluding that the student is indeed more competent at
applied than pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

Now consider table~\ref{fisherparkrun}, giving a summary of whether
nonfocal ocmpetitors place ahead of the author, split by venue.  A
two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
surely an unreasonable conclusion given the author's perceived
performance stagnation.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{parkruncontour} % produced by very_simplified_likelihood.Rmd
\caption{Support contours for Bradley-Terry strength
  for \label{parkruncontour} 2023 and 2024.  Null of equal strengths
  shown as a diagonal line}
\end{centering}
\end{figure}

\begin{table}[h]
    \begin{tabular}{ l c r }
       & \multicolumn{2}{c}{Faster?}\\
       & yes  & no  \\
  2023 & 1506 & 806 \\
  2024 & 2102 & 708
    \end{tabular}
    \caption{Two-way contingency\label{fisherparkrun} table}
\end{table}

@@

\begin{table*}[t]
  \caption{Educational dataset}
\label{educationtable}
\begin{tabular}{llll}
\hline
                   course&category   &rank&class size\\ \hline
      rings and modules  &  pure     & 9  &      12\\
           group theory  &  pure     & 7  &      17\\
               calculus  & applied   & 2  &      23\\
         linear algebra  & applied   & 3  &       9\\
 differential equations  & applied   & 2  &      13\\
               topology  &   pure    & 8  &      14\\
     special relativity  & applied   & 3  &      13\\
        fluid mechanics  & applied   & 4  &      12\\
            Lie algebra  &   pure    & 9  &      15\\
     \hline
\end{tabular}
\end{table*}

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.542$, $\hat{p}_\mathrm{applied}=0.852$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.06$, at
${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units of
support criterion~\citep{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.11$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.042$.

### 2024-09-17 inst/very_simplified_likelihood.Rmd
corrections

```{r focaleducation}
rank <- c( 9, 7, 2, 3,2 ,8 ,5 ,4 ,9 )
class_size <- c(12, 17,23, 9,13,14,13,12,15)
course <- c("rings and modules","group theory","calculus","linear algebra",
"differential equations","topology","special relativity","fluid mechanics","Lie algebra")
category <- c("pure","pure","applied","applied","applied","pure","applied","applied","pure")
data.frame(course,category,rank,class_size)
```

@@

```{r dopurefurst}
Lpure <- f_vec_arn(a,rank[wp],class_size[wp])
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure,ylab='pure strength')
optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp])},c(0.3,0.6),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa])},c(0.3,0.6),maximum=TRUE)
```

@@

```{r,label=fishbish,cache=TRUE}
x <- seq(from=0.1,to=0.99,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2_ed,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

@@

```{r,label=defsupp2}
jj1 <-
optim(
   c(.4,.8),
   function(vec){
      supp2_ed(vec,rank[wp],rank[wa],class_size[wp]-1,class_size[wa]-1)},
   control=list(fnscale = -1)
)
jj1
```

```{r,label=findmaxlike}
jj2 <-
optimize(function(v){supp2_ed(c(v,v),rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
c(0.1,0.9),maximum=TRUE)
jj2
```

```{r,label=shochisw}
pchisq(2*(jj1$value-jj2$objective),lower.tail=FALSE,df=1)
```

>>>>

```{r focaleducation}
rank       <- c( 9,  7,  2, 3,  2,  8,  3,  4,  9)
class_size <- c(12, 17, 23, 9, 13, 14, 13, 12, 15)
course <- c("rings and modules","group theory","calculus","linear algebra",
"differential equations","topology","special relativity","fluid mechanics","Lie algebra")
category <- c("pure","pure","applied","applied","applied","pure","applied","applied","pure")
data.frame(course,category,rank,class_size)
```

@@

```{r dopurefurst}
Lpure <- f_vec_arn(a,rank[wp],class_size[wp])
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure,ylab='pure strength')
optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp])},c(0.3,0.9),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa])},c(0.3,0.9),maximum=TRUE)
```

@@

```{r,label=fishbish,cache=TRUE}
x <- seq(from=0.1,to=0.99,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2_ed,rank[wp],rank[wa],class_size[wp],class_size[wa])
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

@@

```{r,label=defsupp2}
jj1 <-
optim(
   c(.4,.8),
   function(vec){
      supp2_ed(vec,rank[wp],rank[wa],class_size[wp],class_size[wa])},
   control=list(fnscale = -1)
)
jj1
```

```{r,label=findmaxlike}
jj2 <-
optimize(function(v){supp2_ed(c(v,v),rank[wp],rank[wa],class_size[wp],class_size[wa])},
c(0.1,0.9),maximum=TRUE)
jj2
```

```{r,label=shochisw}
jj1$value
jj2$objective
jj1$value-jj2$objective
pchisq(2*(jj1$value-jj2$objective),lower.tail=FALSE,df=1)
```
